---
title: Practical Machine Learning Assignment
author: "Alessandro Camillò"
date: "October 19, 2015"
output: html_document
---



```{r, echo=FALSE}
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(corrplot)

# Initialize the random number generator for reproducubulity
set.seed(1948) 
```

```{r, echo=FALSE}
loadData <- function(url, name) {
  
  dataDir <- "data/"
  
  ## Create the required working folder.
  if (!file.exists(dataDir)) {
    dir.create(dataDir)
  }
  
  fileName <- paste0(dataDir, name)
  
  ##
  ## Download the database if not already present
  ##
  if (!file.exists(fileName)) {
    message("Downloading training database...")
    download.file(url, fileName)
  }
 
  ## A preliminary exploration revelead the data to contain different labels for the missing value.
  ## This variant of the read.csv will automatically replace the many different strings to the 
  ## classical NA value.
  data_data <- read.csv(fileName, na.strings=c("#DIV/0!"," ", "", "NA", "NAs", "NULL"))
}
```

## Read the data
* Read both training and testing instances into the variables: $training_data$ and $testing_data$.
* The function loadData containts some basic data cleaning statements.
```{r cache = TRUE}
training_data <- loadData("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", "pml-training.csv")
testing_data <- loadData("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", "pml-testing.csv")
```

## Clean the data
* The loadData functions replaces the many different labels used for indicating a not available data with NA.
* We remove the columns containing NA  values.
* We remove unwanted columns, useless to the project analysys.

```{r, results='hide'}
## Now let's  remove the columns tcontaining NA  values.
training_data <- training_data[, colSums(is.na(training_data)) == 0] 
testing_data <- testing_data[, colSums(is.na(testing_data)) == 0] 

## Remove unwanted columns.
training_data <- training_data[, !grepl("^X|timestamp|window|user_name", names(training_data)) ]
testing_data <- testing_data[, !grepl("^X|timestamp|window|user_name", names(testing_data)) ]
```

## Create the prediction model

First we split the given training data into two sets, the training dataset $trainingData$ and the validation dataset $testingData$ of approximately **60%** and **40%** of the total size.
```{r, results='hide'}
pivot <- createDataPartition(training_data$classe, p=0.6, list=FALSE )
trainingData  <- training_data[pivot,]
testingData  <- training_data[-pivot,]
```

Second, we create a predictive model using the Random Forst algorithm (RFA). This choice is motivated by the fact that the RFA automatically selects important variables and is robust against highly correlated variables and outliers in general. 
The number of **cross validation is set to 5**

```{r, cache = TRUE}
ctrl  <- trainControl(method = "cv", 5)
modelRFA <- train(classe ~ ., data = trainingData, method = "rf", trControl = ctrl, ntree = 250)
modelRFA
```


In this third step, we assess the performance of the prediction model testing it against the validation data set $testingData$
```{r}
predictRFA <- predict(modelRFA, testingData)
confusionMatrix(testingData$classe, predictRFA)

accuracy <- postResample(predictRFA, testingData$classe)
out_of_sample_error <- 1 - as.numeric(confusionMatrix(testingData$classe, predictRFA)$overall[1])
accuracy
```

```{r, echo= FALSE}
accuracy_percent = as.numeric( round(100 * accuracy, 2))
out_of_sample_error_percent = round(100 * out_of_sample_error, 2)
```


The estimated model accuracy is **`r accuracy_percent[1]`%** with an estimated out of sample error of **`r out_of_sample_error_percent[1]`%**.

In this fourth and final step step the model is eventually validated with the project's testing data set $testing_data$.
```{r}
result <- predict(modelRFA, testing_data[, -length(names(testing_data))])
result
```

## Appendix: Plots and Graphs

```{r}
treeModel <- rpart(classe ~ ., data = training_data, method="class")
prp(treeModel) # fast plot
```
